\section{Conclusion}

This paper presented an attention-enhanced MobileNetV2 framework for plant disease detection on resource-constrained devices. The proposed model integrates the Convolutional Block Attention Module (CBAM) after the final MobileNetV2 feature map and before global average pooling, enabling sequential channel and spatial attention to refine extracted features while preserving a lightweight architecture.

Experiments used 54,306 PlantVillage images across 38 classes. In the consistent saved-model re-evaluation, MobileNetV2 + CBAM achieved 97.17\% accuracy and 97.15\% weighted F1-score, compared with 96.78\% and 96.73\% for MobileNetV2. Xception remained most accurate at 98.30\%. A matched Grad-CAM analysis showed examples in which CBAM broadened lesion-focused activation and corrected a baseline error, but also a shared failure in which both models attended to damaged tissue and predicted the wrong disease.

MobileNetV2 + CBAM contains 4.02 million parameters, requires 0.604 GFLOPs (approximately 0.302 GMACs), and converts to a 4.07 MiB dynamic-range-quantized TFLite file. Quantized accuracy was 96.70\%, a 0.47-percentage-point decrease from the re-evaluated H5 model. On an Intel i7-11800H CPU with TFLite/XNNPACK and eight threads, its median batch-one latency was 45.42 ms (P95: 102.54 ms), essentially equal to MobileNetV2 under the same protocol. These measurements establish a platform-specific compactness result rather than universal real-time readiness.

The conclusions remain limited by controlled-background data, the legacy generator subsampling, 10 audited exact cross-split duplicate groups, and the absence of independently trained seeds. Removing the seven duplicated test images had a negligible effect, but future splits should group duplicate content before partitioning. The study also does not establish robustness to field clutter, illumination, occlusion, symptom severity, or domain shift. No phone, Raspberry Pi, microcontroller, or energy benchmark was performed. Future work should report mean $\pm$ standard deviation across seeds, compare unfreezing depths and CBAM placements, test SE/ECA/coordinate attention and modern lightweight backbones, validate on field datasets, and measure latency, peak memory, and energy on representative target hardware.
